---
title: "Model Comparision"
author: "Sam Abbott"
date: "04/10/2016"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages, echo=TRUE, message=FALSE, include=FALSE}
#install.packages('tidyverse')
library(tidyverse)

# install.packages("devtools")
library(devtools)

# install_github("sbfnk/fitR@fitcourse_2016")
library(fitR)

#install.packages('coda')
library(coda)
```

#Introduction

Slow covergance occurs when:
- If you initialise the MCMC with a init.theta far from the region of high posterior density, the chain might take a long time to reach this region of interest and you will have to burn a lot of iterations.
- If the covariance matrix of the Gaussian proposal is very different from the posterior, this will result in poor mixing and sub-optimal acceptance rates and you will have to run your chain for longer (or thin it

Approach:

- MCMC on determinisitc model and learn from output of the chain to initialise the stochastic model
- deterministic approximates the stochastic model

#Setting the MCMC

- use adaptive MCMC

```{r set up MCMC, echo=TRUE, message=FALSE}
# the fitmodel
data(SEITL_deter)

# wrapper for posterior
my_posteriorTdC <- function(theta) {
    
    my_fitmodel <- SEITL_deter
    my_init.state <- c(S = 279, E = 0, I = 2, T = 3, L = 0, Inc = 0)
    # note that for the SEIT4L model there are 4 state variables for the T
    # compartment my_init.state <- c('S' = 279, 'E' = 0, 'I' = 2, 'T1' = 3, 'T2'
    # = 0, 'T3' = 0, 'T4' = 0, 'L' = 0, 'Inc' = 0)
    
    return(logPosterior(fitmodel = my_fitmodel, theta = theta, init.state = my_init.state, 
        data = FluTdC1971, margLogLike = dTrajObs, log = TRUE))
    
}

# theta to initialise the MCMC
init.theta <- c(R0 = 2, D_lat = 2, D_inf = 2, alpha = 0.8, D_imm = 16, rho = 0.85)

# diagonal elements of the covariance matrix for the Gaussian proposal
proposal.sd <- c(R0 = 1, D_lat = 0.5, D_inf = 0.5, alpha = 0.1, D_imm = 2, rho = 0.1)

# lower and upper limits of each parameter
lower <- c(R0 = 0, D_lat = 0, D_inf = 0, alpha = 0, D_imm = 0, rho = 0)
upper <- c(R0 = Inf, D_lat = Inf, D_inf = Inf, alpha = 1, D_imm = Inf, rho = 1)

# number of iterations for the MCMC
n.iterations <- 5000

# additional parameters for the adaptive MCMC, see ?mcmcMH for more details
adapt.size.start <- 100
adapt.size.cooling <- 0.999
adapt.shape.start <- 200
```

#MCMC parameter tuning

```{r run MCMC, echo=TRUE, message=FALSE}
# run the MCMC
my_mcmc.TdC <- mcmcMH(target = my_posteriorTdC,
                      init.theta = init.theta,
                      proposal.sd = proposal.sd,
                      limits = list(lower = lower,upper = upper),
                      n.iterations = n.iterations,
                      adapt.size.start = adapt.size.start,
                      adapt.size.cooling = adapt.size.cooling,
                      adapt.shape.start = adapt.shape.start)
```

#Short run analysis

```{r short analysis trace MCMC, echo=TRUE, message=FALSE}
# convert to a mcmc object for coda
trace <- mcmc(my_mcmc.TdC$trace)

# compute the acceptance rate
1 - rejectionRate(trace)

# plot the trace
xyplot(trace)

# Let's find a suitable burning:
plotESSBurn(trace)

# Let's create a new trace without the burning
trace.burn <- burnAndThin(trace, burn = 1000)
xyplot(x = trace.burn)

# Let's check the ESS
effectiveSize(trace.burn)

# autocorrelation plot
acfplot(x = trace.burn, lag.max = 60)

# Let's create a thinned trace
trace.burn.thin <- burnAndThin(trace.burn, thin = 20)
xyplot(x = trace.burn.thin)

# Let's check the ESS
effectiveSize(trace.burn.thin)

# new autocorrelation plot
acfplot(x = trace.burn.thin, lag.max = 60)

# The unthinned trace
summary(trace.burn)

# The thinned trace
summary(trace.burn.thin)

#compare
plotPosteriorDensity(list(unthinned = trace.burn, thinned = trace.burn.thin))
```

#Long run analysis

```{r long analysis trace MCMC, echo=TRUE, message=FALSE}
# load mcmc output
data(mcmc_TdC_deter_longRun)

# create mcmc objects for both traces
trace1 <- mcmc(mcmc_SEITL_theta1$trace)
trace2 <- mcmc(mcmc_SEITL_theta2$trace)

# combine traces as mcmc.list object
trace <- mcmc.list(list(trace1, trace2))

# let's have a look
head(trace, 3)


# acceptance rate
1 - rejectionRate(trace)

# close to the optimal value of 0.234

# ESS
effectiveSize(trace)
#due to the high sample size can be very generous in the choice of burnin and thinning

# plot the traces
xyplot(trace)

trace.burn <- burnAndThin(trace, burn = 5000)

# removing the burning increases the ESS
effectiveSize(trace.burn)

# autocorrelation
acfplot(trace.burn, lag.max = 60)

# Thinning: let's keep 1 iteration every 40
trace.burn.thin <- burnAndThin(trace.burn, thin = 40)
xyplot(trace.burn.thin)

# Note that plotPosteriorDensity can take a list of mcmc.list It will plot
# the different mcmc.list by combining their elements Let's plot the
# combined unthinned trace vs the combined thinned trace.
plotPosteriorDensity(list(unthinned = trace.burn, thinned = trace.burn.thin))

#Due to high sample size thinning has almost no effect
#Thinning should only be used for memory or time reasons

densityplot(trace.burn.thin) 

# the function summary combines the chains of a mcmc.list
summary(trace.burn.thin)

#Prior to this amalgimating step must check that the posterior is not multi model

#assess fit of the model
# load data
data(FluTdC1971)

# the same init.state as for the fit
init.state <- c(S = 279, E = 0, I = 2, T = 3, L = 0, Inc = 0)

# by default plotPosteriorFit summarize the fit of 100 thetas sampled from
# the posterior
plotPosteriorFit(trace = trace, fitmodel = SEITL_deter, init.state = init.state, 
    data = FluTdC1971)

# alternatively, one can plot the fit of the mean of the posterior (in this
# case the observation is replicated 100 times)
plotPosteriorFit(trace = trace, fitmodel = SEITL_deter, init.state = init.state, data = FluTdC1971, posterior.summary = "mean")

# or using the maximum a posteriori (MAP) estimate
plotPosteriorFit(trace = trace, fitmodel = SEITL_deter, init.state = init.state, data = FluTdC1971, posterior.summary = "max")
```

#Correlations
```{r correlations MCMC, echo=TRUE, message=FALSE}
# levelplot doesn't accept `mcmc.list`, we pass the first `mcmc` only.
levelplot(trace.burn.thin[[1]], col.regions = heat.colors(100))

with(as.data.frame(trace.burn.thin[[1]]), quantile(R0/D_inf, probs = c(0.025, 0.25, 0.5, 0.75, 0.975)))

#uniform priors means that values are unrealistic in this instance
```

- modify priors with prior information for improved realism

```{r modify priors MCMC, echo=TRUE, message=FALSE}
SEITL_deter$dprior <- function(theta) {
    
    # package with truncated normal distribution
    library(truncnorm)
    
    log.prior.R0 <- dunif(theta[["R0"]], min = 1, max = 50, log = TRUE)
    # normal distribution with mean = 2 and sd = 1 and truncated at 0
    log.prior.latent.period <- log(dtruncnorm(theta[["D_lat"]], a = 0, b = Inf, 
        mean = 2, sd = 1))
    # normal distribution with mean = 2 and sd = 1 and truncated at 0
    log.prior.infectious.period <- log(dtruncnorm(theta[["D_inf"]], a = 0, b = Inf, 
        mean = 2, sd = 1))
    log.prior.temporary.immune.period <- dunif(theta[["D_imm"]], min = 0, max = 50, 
        log = TRUE)
    log.prior.probability.long.term.immunity <- dunif(theta[["alpha"]], min = 0, 
        max = 1, log = TRUE)
    log.prior.reporting.rate <- dunif(theta[["rho"]], min = 0, max = 1, log = TRUE)
    
    return(log.prior.R0 + log.prior.latent.period + log.prior.infectious.period + 
        log.prior.temporary.immune.period + log.prior.probability.long.term.immunity + 
        log.prior.reporting.rate)
    
}
```

# Informative priors

```{r info priors MCMC, echo=TRUE, message=FALSE}
# create mcmc object
trace.info1 <- mcmc(mcmc_SEITL_infoPrior_theta1$trace)
trace.info2 <- mcmc(mcmc_SEITL_infoPrior_theta2$trace)

# combine in a mcmc.list
trace.info <- mcmc.list(trace.info1, trace.info2)

# burn and thin as the chain with uniform prior (see above sections)
trace.info.burn.thin <- burnAndThin(trace.info, burn = 5000, thin = 40)

# check that both chains converged to the same posterior
plotPosteriorDensity(trace.info.burn.thin)

# compare the effect of informative priors on the posterior distribution
plotPosteriorDensity(list(unif = trace.burn.thin, info = trace.info.burn.thin))
```

#Model Comparision

```{r compare DIC, echo=TRUE, message=FALSE}

# combine the two chains in a data frame
trace.combined <- ldply(trace.info.burn.thin)

# take the mean of theta
theta.bar <- colMeans(trace.combined[SEITL_deter$theta.names])
print(theta.bar)

# compute its log-likelihood
init.state <- c(S = 279, E = 0, I = 2, T = 3, L = 0, Inc = 0)
log.like.theta.bar <- dTrajObs(SEITL_deter, theta.bar, init.state, data = FluTdC1971, log = TRUE)
print(log.like.theta.bar)

# and its deviance
D.theta.bar <- -2 * log.like.theta.bar
print(D.theta.bar)

# the effective number of parameters
p.D <- var(-2 * trace.combined$log.likelihood)/2
print(p.D)

# and finally the DIC
DIC <- D.theta.bar + 2 * p.D
print(DIC)

```

# Posterior predicitve P checks

```{r max obs in data, echo=TRUE, message=FALSE}
# This is a function that takes 4 arguments:
# - trace, a data frame containing samples from the posterior
#   distribution, one column per parameter 
# - n.samples, the number of samples to take
# - fitmodel, the model we use to generate replicates
# - init.state, the initial state
# - data, the data set we have fit the model to
# It returns the two-sided p-value for the maximal observation
# in the data with respect to the model.
my_postPredCheck <- function(trace, n.samples, fitmodel, init.state, data) {

    # calculate maximum in obs column of data
    max.data <- max(data$obs)
    
    # draw n.samples random numbers between 1
    # and n.samples using the `samples` function 
    samples <- sample(seq_len(nrow(trace)), n.samples)

    # initialise vector of model maxima
    max.model <- c()
    
    # loop over samples
    for (i in samples) {

        # get i'th column from the trace, unlist
        # (to convert to a vector) and assign to parameter
        # vector theta
        theta <- unlist(trace[i, ])

        # use rObsTraj to generate
        # observation trajectory using theta
        obs.traj <- rTrajObs(fitmodel, theta, init.state, data$time)

        # calculate maximum in model and add to max.model vector
        max.model <- c(max.model, max(obs.traj$obs))
    }

    # calculate quantiles of model maxima
    max.model.quant <- quantile(max.model, probs = c(0.025, 0.975))

    # calculate 2-sided p-value,
    # that is the proportion of elements of max.model which are
    # either greater or equal or less or equal (whichever is
    # less) and  multiply by 2 (because it is a 2-sided test)
    pvalue <- min(sum(max.model <= max.data),
                  sum(max.model >= max.data)) / n.samples * 2
    
    # return two-sided p-value
    return(pvalue)
}

init.state <- c(S = 279, E = 0, I = 2, T = 3, L = 0, Inc = 0)
my_postPredCheck(trace = mcmc_SEITL_theta1$trace[SEITL_deter$theta.names],
                 n.samples = 100,
                 fitmodel = SEITL_deter,
                 init.state = init.state,
                 data = FluTdC1971)
```