---
title: "Modelling Interlude"
author: "Sam Abbott"
date: "03/10/2016"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages, echo=TRUE, message=FALSE, include=FALSE}
#install.packages('tidyverse')
library(tidyverse)

# install.packages("devtools")
library(devtools)

# install_github("sbfnk/fitR@fitcourse_2016")
library(fitR)

#install.packages('coda')
library(coda)
```

#Introduction

[**Lecture slides**](http://sbfnk.github.io/mfiidd/slides/sto_2016.pdf)

```{r load models}
data(SEITL_deter)
data(SEITL_stoch)
```

```{r deterministic simulation, message=FALSE, results='hide', caption='Deterministic simulation of SEITL model with measurement uncertainty and an intial guess of model parameters'}
theta.guess3 <- c(R0 = 10, D_lat = 2, D_inf = 2, alpha = 0.4, D_imm = 13, rho = 0.7)
init.state.bad.guess <- c(S = 250, E = 0, I = 4, T = 0, L = 30, Inc = 0)
plotFit(SEITL_deter, theta.guess3, init.state.bad.guess, data = FluTdC1971)

plotFit(SEITL_deter, theta.guess3, init.state.bad.guess, data = FluTdC1971, n.replicates = 100)
```

```{r stoch simulation single, message=FALSE, results='hide', caption='stochastic simulation of SEITL model with measurement uncertainty and an intial guess of model parameters'}
plotFit(SEITL_stoch, theta.guess3, init.state.bad.guess, data = FluTdC1971, 
    n.replicates = 100)
```

```{r stoch simulation, message=FALSE, results='hide', caption='stochastic simulation of SEITL model with measurement uncertainty and an intial guess of model parameters; Showing all modelled populations'}
plotFit(SEITL_stoch, theta.guess3, init.state.bad.guess, data = FluTdC1971, all.vars = TRUE, n.replicates=100)
```

#Exponential vs Erlanf distributions

- assuming transitions happen at a constant rate implies that the process is memoryless
- to include a memory affect swap expoential with an erlang distribution
- This distribution is parametrised by its mean mm and shape kk and can be modelled by kk consecutive sub-stages, each being exponentially distributed with mean m/km/k.
- memory effect added for the cellular response


```{r load models erlang}
data(SEIT4L_deter)
data(SEIT4L_stoch)
```

```{r det sim: Erlang, message=FALSE, results='hide', caption='Deterministic simulation of SEITL model with measurement uncertainty and an intial guess of model parameters. Erlang distribution for treatment response'}
init.state.guess.SEIT4L <- c(S = 279, E = 0, I = 2, T1 = 3, T2 = 0, T3 = 0, 
    T4 = 0, L = 0, Inc = 0)
plotFit(SEIT4L_stoch, theta.guess3, init.state.guess.SEIT4L, data = FluTdC1971, 
    n.replicates = 100, non.extinct = c("E", "I"))
```

